---
title: "Diabetes Risk Factors Analysis"
author: "Karissa Whiting"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: paper
    toc: yes
    toc_float: yes
fontsize: 12pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(gtsummary)
library(labelled)
library(here)
```

```{r}
load(here("data", "diabetes.RData"))
```

# Methods

Insert methods here....

# Results

```{r}
# function to turn df names to labels
make_labels <- function(names) {
  str_replace_all(names, "_", " ") %>%
    str_to_title() %>%
    return
}

# Prep data -------------
df <- df %>%
  
# make diabetes binary variable (hemoglobin > 7 is considered diabetic)
  mutate(
    diabetes = case_when(
      glyhb >= 7 ~ "Diabetes", 
      glyhb < 7 ~ "No Diabetes", 
      TRUE ~ NA_character_ )) %>% 
    
  # TODO - make variable called diabetes_bin that codes "Diabetes" as 1 and "No Diabetes" as 0
  #  diabetes_bin = ) %>%
  
  #TODO-  make a new variable waist_hip_ratio
  #?????????????? %>% 
  
  # deselect unwanted variables 
  select(-id, -bp_2s, -bp_2d, -bp_1d, -bp_1s,
         -time_ppn, -waist, -hip)


```

- This data set contains `r nrow(df)` observations of individuals with data on risk factors for diabetes. 
- There are `r nrow(df %>% filter(diabetes == "Diabetes"))` individuals with diabetes and `r nrow(df %>% filter(diabetes == "No Diabetes"))` individuals without diabetes in this data set. 

<br>

```{r }
# table 1 of patient characteristics
df %>%
  set_variable_labels(.labels = make_labels(names(.))) %>%
  filter(!is.na(diabetes)) %>%
  tbl_summary(by = "diabetes") %>%
  add_p() %>%
  tab_style_bold_labels() %>%
  tab_style_italicize_levels() %>%
  as_gt() %>%
  tab_options(
    table.font.size = "small",
    row.padding = gt::px(1)) %>%
  tab_header(title = "Patient Characteristics By Diabetes Status")
  
```

<br> 

```{r }

# plot distributions of numeric variables
df %>%
  select(diabetes, age, chol, hdl, height, 
            ratio_chol_hdl, 
            stabilized_glucose, waist_hip_ratio, 
            weight) %>%
  gather(key, value, -diabetes) %>%
  ggplot(aes(x = value, fill = diabetes), alpha = .5) + 
  geom_histogram()  +
  facet_wrap(~key, scales = "free") + theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ggtitle("Continuous Variables by Diabetes Status")
```

<br>

```{r}

# plot univariate regressions continuous 
df %>%
  select(age, chol, hdl, height, 
            ratio_chol_hdl, 
            stabilized_glucose, waist_hip_ratio, 
            weight, glyhb)  %>%
  gather(key, value, -"glyhb") %>%
  ggplot(aes(x = value, y = glyhb, color = key)) + 
  geom_point(alpha = .3, size = .8)  +
  geom_smooth(method='lm') +
  facet_wrap(~key, scales = "free") + theme_minimal() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 90, hjust = 1)) + 
  ggtitle("Continuous Variables by Hemoglobin Levels")

```

<br>

```{r}
# plot outcome by factor 
df %>%
  select(gender, location, frame, glyhb) %>%
  gather(key, value, -"glyhb") %>%
  ggplot(aes(x = value, y = glyhb, fill = key)) + 
  geom_boxplot(alpha = .3, size = .8)  +
  facet_wrap(~key, scales = "free") + theme_minimal() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 90, hjust = 1))  + 
  ggtitle("Categorical Variables by Hemoglobin Levels")
```

<br>

## Linear Model 

### Univariate Regressions

```{r}
# univariate regressions

df %>%
  set_variable_labels(.labels = make_labels(names(.))) %>%
  select(-diabetes, -diabetes_bin) %>%
  tbl_uvregression(method = lm, 
                   y = glyhb) %>%
  tab_style_bold_labels() %>%
  tab_style_italicize_levels() %>%
  as_gt() %>%
  tab_options(
    table.font.size = "small",
    row.padding = gt::px(1)) %>%
  tab_header(title = "Univariate Regression Coefficients")
  
```

### Multivariate Regressions

```{r}
# multivariate model

# set labels for table display
df <- df %>%
  set_variable_labels(.labels = make_labels(names(.))) 

# build the model
multi_lm <- lm(glyhb ~ ratio_chol_hdl +
                 stabilized_glucose +
                 age + weight + waist_hip_ratio, data = df)

#tidy(multi_lm)

# display tidy model results
multi_lm %>%
  tbl_regression() %>%
  tab_style_bold_labels() %>%
  tab_style_italicize_levels() %>%
  as_gt() %>%
  tab_options(
    table.font.size = "small",
    row.padding = gt::px(1)) %>%
  tab_header(title = "Multivariate Model")

```


<br> 

```{r}
pred <- predict(multi_lm)

obs <- df %>%
  select(glyhb, ratio_chol_hdl, stabilized_glucose, 
         age, weight, waist_hip_ratio) %>%
  na.omit() %>% # because model will omit observations that have na
  pull(glyhb)

mod_results <- bind_cols(predicted = pred, 
                         observed = obs)

ggplot(mod_results, aes(x = observed, y = predicted)) +
  geom_point(color = "skyblue3") +
  theme_minimal() +
  ggtitle("Observed vs. Fitted Glyhb Values")
```

<br>

## Logistic Model

### Univariate Regressions

```{r }

# univariate coefficients - exponentiate to get OR's
df %>%
  set_variable_labels(.labels = make_labels(names(.))) %>%
  select(-glyhb, -diabetes) %>%
  tbl_uvregression(method = glm, 
                   y = diabetes_bin, 
                   method.args = list(family = binomial), 
                   exponentiate = TRUE) %>%
  tab_style_bold_labels() %>%
  tab_style_italicize_levels() %>%
  as_gt() %>%
  tab_options(
    table.font.size = "small",
    row.padding = gt::px(1)) %>%
  tab_header(title = "Univariate Regression Coefficients")
  
```


### Multivariate Regressions

```{r}
# multivariate model

# create labels for reporting table
df <- df %>%
  set_variable_labels(.labels = make_labels(names(.))) 

# build model
multi_lm <- glm(diabetes_bin ~ ratio_chol_hdl +
                 stabilized_glucose +
                 age + weight + waist_hip_ratio, data = df, 
                family = binomial)


# display tidy model results
multi_lm %>%
  tbl_regression(exponentiate = TRUE) %>%
  tab_style_bold_labels() %>%
  tab_style_italicize_levels() %>%
  as_gt() %>%
  tab_options(
    table.font.size = "small",
    row.padding = gt::px(1)) %>%
  tab_header(title = "Multivariate Model")

```

